---
title: "ActivityTrack"
output: html_document
---
This assignment will be described in multiple parts. You will need to write a report that answers the questions detailed below.

1. Loading and preprocessing the data
2. What is mean total number of steps taken per day?
3. What is the average daily activity pattern?
4. Imputing missing values

### Loading and preprocessing the data
1.1 Using readCSV() load the data. 


```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  if (!require("knitr")) {
   install.packages("knitr")
  }
  
  if (!require("utils")) {
   install.packages("utils")
  }
  require("knitr")
  require("utils")
  
  df = read.csv(file="C:/Git/Mod_5_Ass_1/activity.csv", header=TRUE, sep=",")
```

1.2 Before we process the data, let us examine the data. Get a data summary

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  summary(df)
```

1.3 Display the internal structure of df

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  str(df)
```

###What is mean total number of steps taken per day?
2.1 Using aggregate, get the sum of steps in a given day

```{r, echo=TRUE, warning=FALSE, message=FALSE}   


 if (!require("stats")) {
   install.packages("stats")
  }
  require("stats")
 #total number of steps per date
 mdata <- aggregate(steps ~ date, data = df, sum)

 head(mdata)

 
```
2.1.1 Get a Summary of the data

```{r, echo=TRUE, warning=FALSE, message=FALSE}   
#summary(mdata)

```

2.2 Make a histogram of the total number of steps taken each day


```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  
  if (!require("ggplot2")) {
    install.packages("ggplot2")
  }
  
  require("ggplot2")
  #plot data
  a <- ggplot(data = mdata, aes(steps))
  a <- a + xlab("Steps") + ggtitle("Histogram of the total number of steps taken each day") + geom_histogram()
  
  print(a)
  

 
```

###Calculate and report the mean and median of the total number of steps taken per day
3.1 Calculate the Mean of the total number of Steps Per day

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  mean(mdata$steps)
```
3.2 Calculate the Median of the total number of Steps Per day

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  median(mdata$steps)
 
```

###What is the average daily activity pattern?

4.1 Calculate the average number of steps for each interval.

```{r, echo=TRUE, warning=FALSE, message=FALSE}   
  #calculate average number of steps per interval
  avstepsperinterval <- aggregate(steps ~ interval, data = df, mean)
 
```

4.2 Plot the time series graph.

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  
 if (!require("ggplot2")) {
    install.packages("ggplot2")
  }
  #plot data
  require("ggplot2")
  a <- ggplot(data=avstepsperinterval, aes(x = strptime(sprintf("%04d", interval), format="%H%M"),steps))
  a <- a + geom_line() +  scale_color_manual(values=c("#CC6666", "#9999CC"))
  a <- a + xlab("Interval Time Slice") + ylab("Average Number of Steps") + ggtitle("Average daily activity pattern") 
  print(a)
 
```

4.3 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

4.3.1 Add the timesliced to it's own vector.

```{r, echo=TRUE, warning=FALSE, message=FALSE}   
  
  #create the timeline interval
  timeslicedInterval<-strptime(sprintf("%04d", avstepsperinterval$interval), format="%H%M")
  head(timeslicedInterval)
 
```

4.3.2 Add the vector to the **avstepsperinterval** variable.

```{r, echo=TRUE, warning=FALSE, message=FALSE}   
    
   #add the time sliced data to the avstepsperinterval variable
   avstepsperinterval<-cbind(avstepsperinterval,timeslicedInterval)

 
```


4.3.3 Using Summary get the Max number of steps.

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  
  summary(avstepsperinterval)
 
```

*Max.   :2355.0   Max.   :206.170   Max.   :2015-07-12 23:55:00*

###Imputing missing values?

5.1 Calculate and report the total number of missing values in the dataset?


```{r, echo=TRUE, warning=FALSE, message=FALSE}   
  #count the number of NA's 
  sum(is.na(df$steps))
 
```
5.2 Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. 

When I looked at this I did 2 strategies, 
1. I replaced NA's with 0.
2. I replaced NA's with the mean of the Interval.
*Strategy 1- Replace NA's with 0*

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  imputedmean<-df
  #replace NA's with 0
  imputedmean[is.na(imputedmean)] <- 0
   
```

5.2.1 Inspect the new dataset **imputedmean**. 


```{r, echo=TRUE, warning=FALSE, message=FALSE}   
  #inspect the data
  head(imputedmean)

  summary(imputedmean)
 
```

5.3 Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

5.3.1 Using aggregate calculate the total number of steps each day

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  #Get the total number of steps
  totalstepsperdays <- aggregate(steps ~ date, data = imputedmean, sum)
 
```
5.3.2 Inspect the data

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  head(totalstepsperdays)
  
  summary(totalstepsperdays)
 
```

5.3.2 Make a histogram of the total number of steps taken each day.

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

 if (!require("ggplot2")) {
    install.packages("ggplot2")
  }
  
  require("ggplot2")
  #plot the histogram
  a <- ggplot(data = totalstepsperdays, aes(steps))
  a <- a + xlab("Steps") + ggtitle("Histogram of the total number of steps taken each day") + geom_histogram()
  
  print(a)
 
```

5.4 What would the histogram look like if I used the mean of the intervals. 
*Strategy 2 -Replace  NA's with the mean of the interval*

5.4.1 Replace NA's with the Mean. 


```{r, echo=TRUE, warning=FALSE, message=FALSE}   
  
  #replace NA's with mean
  imputedmean2<-df
  for(i in 1:nrow(imputedmean2)) {
    
    if(is.na(imputedmean2$steps[i]))
      #replace NA's with mean in avstepsperinterval$steps
      imputedmean2$steps[i] <- avstepsperinterval$steps[i]
  }

  head(imputedmean2)

```

5.4.2 Get a new total using imputed data frame with mean. 


```{r, echo=TRUE, warning=FALSE, message=FALSE}   
  #Get a new total using imputed data frame with mean
  totalstepsperdays_with_mean_data <- aggregate(steps ~ date, data = imputedmean2, sum)
  head(totalstepsperdays_with_mean_data)

```

5.4.3 Graph the imputed data with the mean.
```{r, echo=TRUE, warning=FALSE, message=FALSE}   
 if (!require("ggplot2")) {
    install.packages("ggplot2")
  }
  
  require("ggplot2")
  
  #plot the histogram
  a <- ggplot(data = totalstepsperdays_with_mean_data, aes(steps))
  a <- a + xlab("Steps") + ggtitle("Histogram of the total number of steps taken each day") + geom_histogram()
  
  print(a)
```

5.3.3 Do these values differ from the estimates from the first part of the assignment?

This was interesting, because this depends on what strategy you use. Orinaly I used the strategy where all NA's were replaced by 0. I did this, thinking it would have no affect on calculations. However, there was a change in the histogram, because now we have an increase in 0's. 

I decided to look at the suggestion that was presented in the question, I chose the strategy of assigning all NA's the mean of the interval for the day. The graphs were the same. 

###Are there differences in activity patterns between weekdays and weekends?

6.1 Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day?


6.1.1 Identify weekends.

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  if (!require("timeDate")) {
    install.packages("timeDate")
  }
  
  require("timeDate")

  #replace NA's with mean
  fulldata<-df
  for(i in 1:nrow(fulldata)) {
    
    if(is.na(fulldata$steps[i])){
      #replace NA's with mean in avstepsperinterval$steps
      fulldata$steps[i]<- avstepsperinterval$steps[i]
    }
  }
  head(fulldata)
  fulldata$isaweekend <-isWeekday(as.POSIXct(fulldata$date))
  
 #replace true and false with weekend or weekday
 for(i in 1:nrow( fulldata)) {
    
    if(fulldata$isaweekend[i] ==FALSE){
        #replace FALSE with weekday
        fulldata$isaweekend[i] <- "weekday"
      }else{
         #replace TRUE with weekend
        fulldata$isaweekend[i] <- "weekend"
      }
  }
 
 
 
  
```
6.1.2 Find the mean steps per interval and weekday/weekend.

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  
  weekday_and_weekend_ave <-
    aggregate(steps~ interval+ isaweekend, data=fulldata,FUN=mean)

  weekday_and_weekend_ave$isaweekend <- as.factor(weekday_and_weekend_ave$isaweekend)
  head(weekday_and_weekend_ave)

  
```
6.1.3 Plot Graph.

```{r, echo=TRUE, warning=FALSE, message=FALSE}   

  if (!require("ggplot2")) {
   install.packages("ggplot2")
  }
  head(weekday_and_weekend_ave)
  require("ggplot2")  
  a <- ggplot(data=weekday_and_weekend_ave, aes(x = interval,
                                                y = steps,
                                                colour=isaweekend, 
                                                group=isaweekend))
  a <- a + geom_line() 
  a <- a +facet_grid(.~isaweekend)
  a <- a + xlab("Interval") + ylab("Steps") + ggtitle("Weekend Versus Weekday Activity") 
  print(a)  
  
```